---
title: "Data Wrangling in R"
output: html_notebook
---

Today we are going to learn about data wrangling.  What does it mean to "wrangle data"?  Simply put, wrangling your data is formatting and arranging it for easy and efficient manipulation.

Wangling (or munging, or manipulation) in R primarily uses the `dplyr` and `tidyr` packages.  These packages are core elements of the `tidyverse`, which is RStudio's modern methodology for data wrangling.  So, while you can certainly load all the core packages using `tidyverse`, we will be focusing on just `dplyr` and `tidyr`.

```{r libraries}
library(dplyr)
library(tidyr)
```


## First Steps

The packages make use of a new type of data frame called, `tbl_df` aka tibbles (sounds a lot like tables).  Tibbles are mostly just a pretty way to view large (wide and/or long) data, but they form the foundation of modern R.  Realistically, you as a user won't notice the differnce because access to the data remains consistent with base R data frames.  You should be aware of them and the tools.

The first thing to do with a new data set is understand it's structure and format.  How many columns does it have?  How many rows?  The `View()` (with a Capital V!) opens a spreadsheet-like view of your data to quickly visualize and filter the results.

```{r use-View}
# view the built-in mtcars data using View().
# This will open the data set in a spreadsheet-like view.
View(mtcars)
```


The `tidyverse` introduces a new operator, called a **pipe** to the R lexicon. Pipes, `%>%`, are used for "chaining" results together.  Chaining is a method for stringing together operations, one after another.  Another way to think of the syntax is the form: use this data to do this to it, then do this other thing, then do this next thing, and so on.

Pipes pass the result of the previous operation (function) invisibly to the next operation.  You can call the data explicitly as a named argument, `.data`, but that is not necessary.

To introduce this, let's first look at the standard line-by-line variable assignment.

```{r rounded-mean-line}
# singular, line-by-line method to calculate the mean and round a vector of data
#  will will use mtcars, which is a built in dataset of car attributes to play with

# first, extract the vector (column) of data from the data frame
mpg_v <- mtcars$mpg

# What does the data look like?
mpg_v

# next, calculate the mean
mpg_m <- mean(mpg_v)
mpg_m

# now round the result to two decimal points
round(mpg_m, 2)

```

The same can be accomplished using the pipe method.

```{r rounded-mean-pipe}
# using a piped method
mtcars$mpg %>% #use this data
  mean() %>% #then calculate the mean
  round(2) # then round the result

# ---------------
# we can also assign the result to a variable
# mpg_m_p <- mtcars$mpg %>% #use this data
#   mean() %>% #then calculate the mean
#   round(2) # then round the result
# print the result
# mpg_m_p
```

The example above did not keep the variables like the line-by-line method, but we could very easliy have assigned the result to a variable.  The example above is a very simple example, and the benefits of piping might not be immediately obvious.  However, piping in more complex analysis clarifies the intent of your code and makes for readable code.


## Tidy data

Tidy data is organized data that is efficient for R to work with.  Each *variable* in our data is a **column** and each *observation* is in a **row**.  Remember, R is vectorized, meaning our variables are vectors of data over which we can quickly and efficiently perform operations.

Untidy data is a format that does not follow the variable/observation format.  It's not that the untidy data cannot be used in R.  Rather, the format makes indexing the variables more difficult, and more complex code is required to make this work.

Consider the following example:

```{r untidy-data}
# we can create a dummy data set to illustrate untidy data.
untidy_df <- tibble(country = c("US", "CA", "DE"),
                    `2010` = c(7000, 6500, 5000),
                    `2015` = c(3000, 4500, 6000),
                    `2017` = c(1500, 2000, 7000)
                    )
untidy_df
```

These data are in a format that does not allow for easy manipulation of the variables.  This format is often referred to as *wide*, where R is looking for *long* data.  In this case, we cannot, for example, quickly extract all years using a single variable, `year`, for use as a grouping variable for summary statistics.

This is not an uncommon format for data.  In fact, the format above is common when *collecting* data, but is not as easily used for *analysis* of those data.

The `tidyr` package offers several functions for working with these untidy data.  We can `gather()` the data into the format that R prefers.  We might want three columns with `country`, `year`, `n`.  We do this by using **key**, **value** pairs.  A **key** column contains the values from former column names, and a value column contains the **values** from each of those columns.

```{r gather}
# gather the untidy data into a tidy format.
# The country column will be maintained and will be extended as necessary
# The format of the key/value arguments are the word key = "new_column_name_to_use"

tidy_df <- gather(untidy_df, # the data to use
                  key = "year", # the key (column names) that will become the categorical or factor
                  value = "n", # the values that are associated with each key
                  2:4) #the columns to use, skip the first column, country.  
tidy_df
```


Notice that the country column values are replicated as needed to create observations.  This might seem redundant compared to the untidy example, but you will see how this tidy data is really powerful.

## The `dplyr` Package

We can now use this tidy data to find specific answers to questions.  `dplyr` has a core set of functions that will be most helpful.  The full set of functions can be found in the (Data Transformation Cheatsheet)[https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf].

For fun we will use the new `starwars` data set that comes with `dplyr`

```{r startwars}
# look at the star wars data
starwars

```


### `Select()`

The `select()` function simplifies the extraction of columns (variables) of data.  Base R certainly has the ability to extract columns of data.  For example:

```{r base_r}
# select several columns by name
starwars[c("name", "height", "skin_color")]

# select by position
starwars[c(1,2,5)]
```

Now try using `select()` from `dplyr`.  The code below uses pipes, but the functions also work without them.

```{r select}
# select just a few columns
starwars %>% #the data set
  select(name, height, skin_color) #the column names

# same call without the pipe
# select(.data = starwars, name, height, skin_color)
```

We can also use a range of values

```{r select-range}
# now try it using a range selection to get the same set
starwars %>% 
  select(1:2,5)
```

The primary difference in using `select()` is the simplified code (no [] or "" or c()), and the addition of "helper" functions used the slice and dice your data down to extactly what you are interested in.

We can use negation to exclude columns, keeping all the rest.

```{r select-inverse}
# select by inverse
starwars %>% 
  select(-name, -height, -skin_color)
```

We can extract the columns that start with a particular letter of the alphabet or search for a partial string.

```{r select-string}
# select just the columns that start with h
starwars %>% 
  select(starts_with("h"))  
```

There are other helper functions that I'm not covering for simplicity, but you are encouraged to look at the cheatsheet or the website documentation for more examples.


### `filter()`

The `filter()` function simplifies the extraction of rows (observations) meeting some criteria.

Again, the same can be done in base R.

```{r base-r-filter}
# the first three observations (rows)
starwars[1:3,]
```

```{r base-r-filter2}
# a more complex conditional example
starwars[which(starwars$height > 160 & starwars$mass < 60), ] #don't forget that last comma!
```

Filtering observations based on specific criteria is where base R becomes a little tedious with all the extra calls to the data set and $ notation.  Also, you have to remember that extra "," at the end of the line to capture all the columns.

The `dplyr` method offers a clear syntax to accomplish the same results.

```{r filter}
# filter the same critera
starwars %>% 
  filter(height > 160 & mass < 60)

```

`filter()` also has helper functions like is.na().

```{r filter-na}
# filter empty observations along with some other criteria
starwars %>% 
  filter(!is.na(mass) & eye_color == "blue" & species != "Human")
```


### `mutate()`

`mutate()` is used to generate new variables.  We can generate multiple values in a single call to the function.  `mutate()` is a window functions that will always return a result of the same length as the input.

```{r bmi}
# calculate the body mass index and convert the height to inches for the starwars dataset
starwars %>% 
  mutate(bmi = height/mass, height_in = height * 0.393701) %>% # pipe the result to a select to reduce the columns
  select(name, height, mass, bmi, height_in)
```

### `summarize()`

`summarize()` will calculate summary statistics.  Unlike `mutate()`, `summarize()` results in a singular value.  All you basic stats are available: mean, median, standard deviation, and other values like counts and sums.

```{r summarize}
# calculate the avg mass and height
starwars %>% 
  summarise(avg_mass = mean(mass, na.rm = TRUE), avg_height = mean(height, na.rm = TRUE))
# notice we had to use the extra argument na.rm = TRUE to exclude NA values for the calculation.
# we get an empty result otherwise.
```


### `arrange()`

Sorting data is done using `arrange()`.  The default is to sort ascending (lowest to highest)  Add the helper function desc() to reverse the order desc() = descending (highest to lowest).  We can also do multiple columns (sort this, then by that).

```{r sort}
# sort by increasing height
starwars %>% 
  arrange(height)

# sort descending
starwars %>% 
  arrange(desc(height))
```

```{r sort-multiple}
# sort by decreasing height, then mass
starwars %>% 
  arrange(desc(height, mass))
# you have to scroll down a little to see the effect.  The second variable becomes the tie-breaker when the values are the same.
```


### `group_by()`

The value of tidy data becomes more apparent when groups come into play.  We can use various variables to group and summarize results.  Using our original tidy data set from above, we can calculate the mean value for each year.

```{r group}
# group by year and calculate the mean, the sum of values, and the number of observations
tidy_df %>% 
  group_by(country) %>% 
  summarise(mean = mean(n), sum = sum(n), n = n())
```

You can also use multiple grouping factors to derive statistics for each unique group.

```{r multiple-groups}
# determine the average hight and mass for each species by gender
starwars %>% 
  group_by(species, gender) %>% 
  summarize(avg_mass = mean(mass, na.rm = TRUE), avg_height = mean(height, na.rm = TRUE)) 
```

## Put it all together

The real power of data wrangling is to find only the data you interested in.  We can put all these methods together into a piped operation to drill down into the data and find answers.

```{r}
starwars %>% #the data
  filter(species %in% c("Human", "Kaminoan", "Twi'lek")) %>% # subset the data to a few species
  group_by(species, gender) %>% # group by species and gender
  summarize(avg_mass = mean(mass, na.rm = TRUE), avg_height = mean(height, na.rm = TRUE)) 
  # calculate the avg mass and height

```


